% !TEX TS-program = xelatex
% Compile with XeLaTeX or LuaLaTeX (because of fontspec)

\documentclass[10pt,oneside]{article}

% ------------------------------------------------
% Layout
% ------------------------------------------------
\usepackage[a4paper,margin=1in]{geometry}

% ------------------------------------------------
% Fonts (XeLaTeX/LuaLaTeX)
% ------------------------------------------------
\usepackage{fontspec}
\setmainfont{Carlito}

% ------------------------------------------------
% Typography & spacing
% ------------------------------------------------
\usepackage{microtype}
\usepackage{setspace}
\onehalfspacing
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.6em}

% ------------------------------------------------
% Math
% ------------------------------------------------
\usepackage{amsmath,amssymb,mathtools}

% ------------------------------------------------
% Tables & lists
% ------------------------------------------------
\usepackage{booktabs}
\usepackage{enumitem}

% ------------------------------------------------
% Hyperlinks
% ------------------------------------------------
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]{hyperref}

% ------------------------------------------------
% Title
% ------------------------------------------------
\title{\textbf{Risk Measurement Exercises: Value-at-Risk (VaR) and Expected Shortfall (ES)}}
\author{}
\date{}

% ------------------------------------------------
% Convenience macros (optional)
% ------------------------------------------------
\newcommand{\VaR}{\mathrm{VaR}}
\newcommand{\ES}{\mathrm{ES}}
\newcommand{\Loss}{L}
\newcommand{\R}{R}
\newcommand{\Normal}{\mathcal{N}}


\title{\textbf{Applied Case Studies}}
\author{Course: Enterprise Risk Management}
\date{}

\begin{document}
\maketitle

\onehalfspacing


\section{Case Study 1: Market Risk, Value-at-Risk and Risk Appetite}

\subsection{Context}

A mid-sized bank operates a trading desk managing a diversified market-risk
portfolio with a total market value of \textbf{€100 million}. The bank has
implemented an ERM framework centered on quantitative risk limits, with
Value-at-Risk (VaR) as the primary risk metric.

The desk is considered compliant as long as its estimated VaR remains below
the approved limit.

\subsection{Portfolio Composition}

The portfolio consists of the following exposures:
\begin{itemize}
    \item Equity exposure: €60 million invested in a broad equity index.
    \item Interest rate exposure: €40 million in a 10-year government bond,
    managed through duration exposure.
\end{itemize}

\subsection{Risk Parameters}

The following risk parameters are provided:
\begin{itemize}
    \item Daily equity volatility: 2\%.
    \item Daily volatility of the 10-year yield: 5 basis points.
    \item Correlation between equity returns and yield changes: 0.
    \item Interest rate sensitivity: a \textbf{+1 bp} increase in yields results
    in a \textbf{€0.20 million loss}.
\end{itemize}

Assume normally distributed risk factors and a one-day holding period.

\subsection{ERM Constraint}

The bank has set the following limit:
\begin{itemize}
    \item 1-day 99\% Value-at-Risk limit: \textbf{€3.0 million}.
\end{itemize}

\subsection{Questions}

\begin{enumerate}
    \item Compute the 1-day 99\% Value-at-Risk of the portfolio using a
    variance--covariance approach.
    \item Determine whether the trading desk complies with the VaR limit.
    \item Consider the following stress scenario:
    \begin{itemize}
        \item Equity index decline of 8\%.
        \item Increase of 30 basis points in the 10-year yield.
    \end{itemize}
    Compute the portfolio loss under this scenario.
    \item Explain why losses of this magnitude may occur despite compliance
    with the VaR limit.
    \item Identify at least two weaknesses of relying exclusively on VaR from
    an ERM perspective.
\end{enumerate}

\section{Case Study 2: Expected Shortfall, Governance and Escalation}

\subsection{Context}

Following internal reviews, the bank complements VaR with Expected Shortfall
(ES) reporting to better capture tail risk. ES is monitored daily and compared
against a predefined limit. Breaches are supposed to trigger immediate
escalation to senior management.

\subsection{Observed Portfolio Performance}

The daily profit-and-loss (P\&L) of the portfolio over the last 12 trading days
(in millions of euros) is reported below:

\begin{center}
\begin{tabular}{cc}
\toprule
Day & P\&L (€m) \\
\midrule
1  & +0.6 \\
2  & -1.1 \\
3  & +0.4 \\
4  & -2.3 \\
5  & +0.9 \\
6  & -0.8 \\
7  & -3.5 \\
8  & +0.5 \\
9  & -1.7 \\
10 & -4.2 \\
11 & -0.6 \\
12 & -6.8 \\
\bottomrule
\end{tabular}
\end{center}

\subsection{ERM Rules}

The ERM framework specifies:
\begin{itemize}
    \item 1-day 99\% Expected Shortfall limit: \textbf{€4.0 million}.
    \item Mandatory escalation in case of a breach.
\end{itemize}

\subsection{Questions}

\begin{enumerate}
    \item Sort the P\&L distribution from worst to best.
    \item Compute the empirical 1-day 99\% Expected Shortfall.
    \item Determine whether the ES limit is breached.
    \item The large loss observed on Day 12 occurred without prior escalation.
    Identify possible failures related to:
    \begin{itemize}
        \item Risk reporting,
        \item Governance architecture,
        \item Incentives or risk culture.
    \end{itemize}
    \item Explain why replacing VaR with Expected Shortfall does not, by itself,
    guarantee effective ERM.
\end{enumerate}



\section{Case study 3 : signals without escalation}

\subsection{Context}

A mid-sized European bank operates across corporate lending, structured finance,
and capital markets activities. The institution is considered well managed and
has experienced several years of stable profitability and moderate growth.

The bank is subject to prudential supervision and regularly undergoes internal
and external reviews of its risk management practices. No material weaknesses
have been identified in recent assessments.

\subsection{Risk governance practices}

Risk management within the bank is organized around a centralized risk function
that operates independently from business units. Risk policies define limits
for key risk categories and specify reporting obligations at different levels of
management.

Senior management receives periodic risk reports summarizing:
\begin{itemize}
    \item[$\bullet$] utilization of risk limits,
    \item[$\bullet$] stress-test outcomes,
    \item[$\bullet$] qualitative risk assessments prepared by risk functions.
\end{itemize}

A committee involving senior executives meets regularly to review these reports
and discuss emerging risks.

\subsection{Strategic development}

Over a two-year period, the bank expanded rapidly in structured finance and
specialized lending activities. Individual transactions complied with existing
risk limits and approval procedures.

However, several features characterized this expansion:
\begin{itemize}
    \item[$\bullet$] repeated concentration in similar counterparties and sectors,
    \item[$\bullet$] increasing reliance on favorable macroeconomic assumptions,
    \item[$\bullet$] frequent use of exceptions and temporary adjustments to
    internal limits.
\end{itemize}

While no formal breaches were recorded, internal risk assessments increasingly
highlighted rising portfolio sensitivity to adverse scenarios.

\subsection{Risk signals}

Risk analysts noted that portfolio-level exposures were approaching internally
defined thresholds under adverse conditions. These observations were included
in written risk reports and discussed during committee meetings.

The language used to describe these risks emphasized uncertainty rather than
immediate concern. Follow-up actions were left to business judgment, and no
explicit escalation was triggered.

\subsection{The event}

A sudden downturn affecting several correlated sectors led to a rapid
deterioration in asset quality and valuation losses. The combined impact
exceeded prior stress-test projections and resulted in a significant reduction
in capital buffers.

Although regulatory minimums were not immediately breached, the bank faced
heightened supervisory scrutiny, rating downgrades, and a loss of market
confidence.

\subsection{Questions}

\begin{enumerate}
    \item Based on the information provided, infer the \textbf{type of ERM
    framework} implemented by the bank. What elements support your inference?

    \item Reconstruct the bank’s \textbf{ERM architecture}. Who appears to:
    \begin{itemize}
        \item[$\bullet$] take risk,
        \item[$\bullet$] monitor and challenge risk,
        \item[$\bullet$] decide on escalation?
    \end{itemize}

    \item Identify the \textbf{early warning signals} that preceded the event.
    Why did they fail to translate into corrective action?

    \item Explain how incentives, governance dynamics, or risk culture may have
    weakened the effectiveness of the ERM system.

    \item Propose two concrete changes that would strengthen the bank’s ERM
    architecture and prevent similar failures.
\end{enumerate}

\section{Case study 4: hedging effectiveness under ERM (prospective vs retrospective)}

\subsection{Context}

A global industrial firm consumes a key commodity input. To stabilize margins,
the firm hedges its input cost exposure using standardized exchange-traded
futures linked to a benchmark commodity index.

Let $X$ denote the firm’s unhedged monthly profit-and-loss (P\&L) exposure
(driven mainly by commodity price changes). Let $H$ denote the monthly P\&L of
the hedging instrument (futures). The firm sets a hedge ratio $h$ and defines
the hedged P\&L as:
\[
Y = X - hH
\]

The goal is to reduce risk (variance and tail losses) \emph{without creating
unacceptable liquidity stress from margin calls}.

\subsection{Data provided}

Risk management estimates hedging parameters using a 24-month historical window
(\emph{normal period}). A market disruption then occurs (\emph{stress period}),
during which correlations and liquidity conditions change.

All statistics below are monthly, expressed in \textbf{€ million}.

\medskip

\begin{center}
\begin{tabular}{lcc}
\hline
 & Normal period (estimation) & Stress period (realized) \\
\hline
$\mathbb{E}[X]$ & 0.0 & -2.0 \\
$\mathbb{E}[H]$ & 0.0 & 0.0 \\
$\mathrm{Var}(X)$ & 100 & 225 \\
$\mathrm{Var}(H)$ & 64 & 144 \\
$\mathrm{Cov}(X,H)$ & 48 & 12 \\
\hline
\end{tabular}
\end{center}

\medskip

In the stress period, the firm also faces exchange margining:
\begin{itemize}
  \item[$\bullet$] Initial margin posted at inception: 10€m (funded from cash).
  \item[$\bullet$] Average monthly \emph{variation margin outflow} on the hedge:
  18€m.
\end{itemize}

Assume the firm has a liquidity buffer of 25€m dedicated to hedging operations.
If monthly margin outflows exceed this buffer, the firm must raise cash via
asset sales or emergency credit.

\subsection{Hedging rules used by the firm}

\begin{itemize}
  \item[$\bullet$] The hedge ratio is set using the minimum-variance benchmark:
  \[
  h^* = \frac{\mathrm{Cov}(X,H)}{\mathrm{Var}(H)}
  \]
  computed from the normal-period estimates.

  \item[$\bullet$] Hedging effectiveness is assessed via variance reduction:
  \[
  \text{HE} = 1 - \frac{\mathrm{Var}(Y)}{\mathrm{Var}(X)}
  \]

  \item[$\bullet$] Risk management reports prospective (ex ante) effectiveness
  using normal-period parameters, then evaluates retrospective (ex post)
  effectiveness using realized stress-period parameters \emph{while keeping the
  same hedge ratio $h$}.
\end{itemize}

\subsection{Tasks}

\begin{enumerate}
  \item \textbf{Hedge ratio.} Compute the minimum-variance hedge ratio $h^*$
  using the normal-period statistics.

  \item \textbf{Prospective effectiveness (ex ante).}
  Using $h=h^*$ and the normal-period statistics, compute:
  \[
  \mathrm{Var}(Y) = \mathrm{Var}(X) + h^2\mathrm{Var}(H) - 2h\,\mathrm{Cov}(X,H)
  \]
  and then compute $\text{HE}_{\text{pros}}$.

  \item \textbf{Retrospective effectiveness (ex post).}
  Using the \emph{same} hedge ratio $h=h^*$ but stress-period statistics,
  compute $\mathrm{Var}(Y)$ and $\text{HE}_{\text{retro}}$.

  \item \textbf{Interpretation.}
  Explain why $\text{HE}_{\text{retro}}$ differs from $\text{HE}_{\text{pros}}$
  in terms of correlation, basis risk, and regime change.

  \item \textbf{Liquidity assessment.}
  Compare the average monthly variation margin outflow (18€m) to the hedging
  liquidity buffer (25€m). Is the buffer sufficient on average?

  \item \textbf{ERM stress question.}
  Suppose variation margin outflows can spike to 40€m in bad months.
  Under ERM, list two governance or design changes that could reduce the
  probability of forced asset sales.

  \item \textbf{ERM integration.}
  Identify at least three risk categories that interacted in this episode and
  explain how ERM should have governed the interactions (not the silos).
\end{enumerate}


\section{Case study 5: when hedging increased risk}

\subsection{Context}

A large industrial firm operates globally and is highly exposed to commodity
price fluctuations. Its core activity relies on a key input whose market price
is volatile and subject to geopolitical and macroeconomic shocks.

To stabilize earnings and protect margins, management has historically relied
on derivative contracts to hedge price risk. The firm is considered
sophisticated and has a dedicated risk management function.

Over recent years, profitability has been stable and volatility of reported
earnings has declined.

\subsection{Risk governance and hedging strategy}

The firm’s risk management policy allows the use of derivative instruments
to hedge commodity price exposure over a rolling horizon.

\begin{itemize}
  \item[$\bullet$] Hedging is conducted primarily using futures and swap
  contracts linked to benchmark commodity prices.

  \item[$\bullet$] Hedge ratios are determined based on historical correlations
  between input costs and output prices.

  \item[$\bullet$] Hedging effectiveness is assessed ex ante using variance
  reduction metrics and reviewed periodically.
\end{itemize}

Risk reports show that the hedging program significantly reduces short-term
price volatility under normal market conditions.

\subsection{Quantitative information}

Let $X$ denote the unhedged price exposure of the firm and $H$ the payoff of
the hedging instrument.

The hedged position is defined as:
\[
Y = X - hH
\]

Based on historical data, the estimated minimum-variance hedge ratio is:
\[
h^* = \frac{\mathrm{Cov}(X,H)}{\mathrm{Var}(H)} = 0.9
\]

Prospective hedging effectiveness is estimated as:
\[
\text{HE}_{\text{pros}} = 1 - \frac{\mathrm{Var}(Y)}{\mathrm{Var}(X)} = 75\%
\]

As a result, management authorizes a large-scale hedging program aligned with
this ratio.

\subsection{The shock}

Following a sudden market disruption:
\begin{itemize}
  \item[$\bullet$] Correlations between the firm’s input prices and benchmark
  contracts weaken significantly.
  \item[$\bullet$] Futures markets experience sharp volatility and liquidity
  deteriorates.
  \item[$\bullet$] Margin requirements on derivative positions increase rapidly.
\end{itemize}

Although the underlying commodity price moves in the expected direction,
the firm faces substantial cash outflows due to margin calls.

\subsection{Outcomes}

Despite being economically hedged in the long term, the firm experiences:
\begin{itemize}
  \item[$\bullet$] severe short-term liquidity stress,
  \item[$\bullet$] forced asset sales to meet margin calls,
  \item[$\bullet$] a deterioration of credit metrics and investor confidence.
\end{itemize}

Retrospective hedging effectiveness falls sharply:
\[
\text{HE}_{\text{retro}} < 0
\]

Senior management questions how a risk-reducing strategy could generate such
outcomes.

\subsection{Questions}

\begin{enumerate}
  \item Identify the main risks faced by the firm. Which risk categories
  interacted in this episode?

  \item From an ERM perspective, explain why the hedging strategy appeared
  effective ex ante.

  \item Interpret the hedge ratio $h^*$. What assumptions underlie its use?

  \item Why did prospective hedging effectiveness fail to predict actual
  outcomes?

  \item Using the ERM framework, identify which secondary risks were
  insufficiently considered.

  \item How should ERM have governed the use of derivatives differently?

  \item Propose two changes to the firm’s ERM architecture or decision process
  that would have reduced the severity of the crisis.
\end{enumerate}

\section{Case study: enterprise risk management at Shell}

Shell plc is a global energy company with operations covering upstream
exploration and production, integrated gas, chemicals, trading, power
generation, and retail activities. Its business model exposes it to a wide
range of uncertainties, including commodity price volatility, geopolitical
risk, operational and safety hazards, regulatory change, and long-term
transition risks related to climate and energy policy. The scale of its
operations, combined with capital intensity and global reach, makes risk
management a central component of corporate governance.

Shell’s current approach to enterprise risk management has been shaped by
historical crises, including major operational incidents and periods of
financial stress. These events revealed weaknesses in governance, escalation,
and oversight, and led to a deliberate strengthening of group-level risk
management. As a result, Shell has positioned ERM not as a purely technical or
compliance-driven activity, but as an integral part of decision-making and
strategic oversight.

At governance level, the Board of Directors retains ultimate responsibility
for risk management and internal control. Specific oversight responsibilities
are delegated to board committees. The Audit and Risk Committee reviews the
effectiveness of Shell’s risk management and internal control systems, while
the Safety, Environment and Sustainability Committee focuses on risks related
to safety performance, environmental impact, and the energy transition. This
structure reflects an explicit recognition that non-financial risks can be as
material as financial risks.

Within management, Shell operates a centralized enterprise risk management
framework. Risk ownership is assigned to senior executives and business
leaders, who remain accountable for the risks arising from their activities.
The central risk function does not own risks; instead, it provides common
methodologies, coordination across the group, and independent challenge. This
separation between ownership and oversight is intended to ensure that risk
remains embedded in line management rather than delegated to specialists.

Shell identifies a limited set of \emph{principal risks} that are disclosed
annually in its reporting. These principal risks represent those that could
materially affect the company’s strategy, performance, financial position, or
reputation. They include, among others, commodity price risk, geopolitical and
regulatory risk, safety and operational incidents, climate transition risk,
cyber risk, and liquidity and funding risk. The list is not static and evolves
as the external environment and strategic priorities change.

Risk assessment combines quantitative analysis with qualitative judgment and
scenario-based thinking. For market-related risks, particularly commodity
price exposure, Shell relies on quantitative metrics, stress testing, and
risk limits within its trading and treasury activities. For longer-term and
structural uncertainties, such as the pace of the energy transition or major
geopolitical shifts, Shell uses scenario analysis rather than point forecasts.
These scenarios are designed to test strategic resilience rather than predict
specific outcomes.

Risk responses are tailored to the nature of each principal risk. Operational
risks are addressed through safety management systems, engineering standards,
and process controls applied across the group. Financial risks, including
commodity price, foreign exchange, and interest rate risks, are managed through
a combination of natural hedging, portfolio diversification, and the use of
derivative instruments. Derivatives are widely used in trading and treasury
activities, but their use is governed by formal policies that define eligible
instruments, limits, approval processes, and escalation thresholds. Hedging is
intended to support the underlying business and protect cash flows rather than
to eliminate all exposure.

Liquidity risk management is treated as a core enterprise risk. Shell maintains
substantial liquidity buffers and diversified funding sources to ensure
resilience under stress. The firm explicitly considers the interaction between
derivative hedging and liquidity, including the potential for margin calls
during periods of market volatility. This reflects an awareness that hedging
strategies can create short-term funding stress even when they are economically
sound over the long term.

Enterprise risk management is embedded in Shell’s strategic planning and
capital allocation processes. Major investment decisions are evaluated not only
on expected returns but also on their robustness under adverse scenarios and
consistency with risk appetite and balance-sheet capacity. Long-term risks,
particularly those related to climate and energy transition, are treated as
strategic uncertainties that cannot be hedged through financial instruments.
Instead, they are managed through portfolio diversification, staged investment,
and strategic flexibility.

Risk reporting is designed to support escalation and decision-making rather
than passive monitoring. Reports to senior management and the board focus on
key risk drivers, changes in the risk profile, and proximity to defined risk
tolerances. When indicators deteriorate or approach limits, management is
expected to respond by adjusting plans, revising exposures, or explicitly
accepting the risk at the appropriate governance level.

Overall, Shell’s enterprise risk management system reflects a principles-based
and integrated approach. Risk management is embedded in governance, strategy,
and capital allocation rather than operating as a separate technical layer.
At the same time, the effectiveness of the system depends heavily on the
quality of assumptions, the willingness of leaders to engage with challenge,
and the organization’s ability to adapt when familiar risk patterns break
down. The case therefore illustrates both how ERM is implemented in practice
by a large multinational firm and why even sophisticated ERM cannot eliminate
the possibility of failure.

\subsection{Discussion questions}

Answer the following questions using evidence from the case above. Answers
should focus on reasoning and diagnosis rather than description.

\subsubsection{ERM framework and philosophy}

\begin{enumerate}
  \item How would you characterize Shell’s ERM framework?
  Is it primarily rules-based or principles-based? Compliance-oriented or
  decision-oriented? Which elements of the case support your view?

  \item Shell focuses on a limited number of \emph{principal risks}.
  What does this reveal about its ERM philosophy, and what are the potential
  trade-offs of this approach?

  \item Why does Shell rely heavily on scenario analysis rather than point
  forecasts for long-term risks? What does this imply about the limits of
  quantitative risk measurement?
\end{enumerate}

\subsubsection{ERM architecture and governance}

\begin{enumerate}
  \setcounter{enumi}{3}
  \item Identify the key components of Shell’s ERM architecture.
  How are risk ownership and risk oversight allocated across the organization?

  \item Why is the separation between business risk ownership and the central
  ERM function emphasized? Under what circumstances could this separation
  weaken risk management?

  \item How might the existence of multiple board-level committees complicate
  coordination and escalation of risks?
\end{enumerate}

\subsubsection{Risk integration and instruments}

\begin{enumerate}
  \setcounter{enumi}{6}
  \item Identify several risk categories addressed in Shell’s ERM system.
  How does the case suggest these risks are considered in an integrated way?

  \item What role do derivative instruments play in Shell’s risk management?
  Why are formal constraints on their use essential from an ERM perspective?

  \item Why does Shell explicitly consider the liquidity implications of
  derivative hedging? What type of failure is this designed to prevent?
\end{enumerate}

\subsubsection{Strategy, limits, and failure modes}

\begin{enumerate}
  \setcounter{enumi}{9}
  \item How is ERM embedded in Shell’s strategic planning and capital allocation
  processes? Give concrete examples from the case.

  \item Which aspects of Shell’s ERM design appear most dependent on managerial
  judgment rather than formal rules? Why does this matter?

  \item Describe a plausible scenario in which Shell’s ERM system could fail
  despite being well designed. What would trigger such a failure, and which
  mechanisms would likely break down first?

  \item Based on the case, explain why effective ERM reduces vulnerability but
  does not eliminate the possibility of failure.
\end{enumerate}

\section{Group exercise: designing ERM for a complex firm}

This exercise asks you to design an enterprise risk management (ERM) system
for a firm operating in a highly uncertain and capital-intensive environment.
You are not asked to replicate an existing framework, but to propose a coherent
ERM design that could realistically function within the organization described
below.

\subsection{Firm description}

Consider a large multinational firm operating in a capital-intensive industry
with global operations. The firm is exposed to volatile input prices, complex
supply chains, geopolitical risk, regulatory uncertainty, and long-term
structural change in its core markets. Its activities generate substantial cash
flows in normal conditions, but downside risks can materialize rapidly and
simultaneously across business lines.

The firm has grown through a combination of organic investment and acquisitions.
As a result, its operations are diversified but organizational complexity is
high. Decision-making authority is largely decentralized, with business units
having significant autonomy over investments and commercial strategy. Central
functions exist, but their influence varies across the organization.

The firm has experienced episodes of stress in the past, including periods of
sharp earnings volatility and operational incidents. These events did not
threaten its survival, but they revealed weaknesses in coordination, escalation,
and anticipation of adverse scenarios. Senior management has therefore decided
to strengthen enterprise-wide risk management.

\subsection{Constraints and objectives}

Senior management has defined the following high-level objectives for ERM:

The ERM system should support strategic decision-making rather than operate as
a compliance or reporting exercise. It should improve the firm’s ability to
anticipate adverse outcomes, manage risk interactions, and preserve resilience
under stress. At the same time, ERM should not paralyze decision-making or
eliminate risk-taking, which remains essential to value creation.

You should assume that management is willing to support ERM in principle, but
that incentives, power structures, and time constraints may limit how far
formal processes can go. Your design should therefore be realistic and
implementable.

\subsection{Tasks}

Your group is asked to propose an ERM design covering the following dimensions.
You should justify each design choice by explicitly linking it to the firm’s
context and constraints.

\paragraph{1. ERM philosophy and scope}

Describe the overall philosophy of your ERM system. Will it be primarily
principles-based or rules-based? Will it focus on a limited set of material
risks or attempt comprehensive coverage? Explain how your chosen philosophy
helps decision-makers under uncertainty.

\paragraph{2. ERM architecture and governance}

Propose an ERM architecture that specifies:
who owns risks,
who oversees risks,
and how risk information flows to senior management and the board.
Explain how your design balances decentralization of decision-making with
enterprise-wide coherence. Identify where challenge should occur and who has
the authority to escalate concerns.

\paragraph{3. Risk identification and assessment}

Explain how material risks should be identified and assessed.
How would you ensure that emerging risks and interactions between risks are
captured, rather than only well-known or easily measurable risks?
Discuss the role of quantitative metrics versus qualitative judgment and
scenario analysis.

\paragraph{4. Risk response and use of instruments}

Describe how your ERM system would govern risk responses.
Which types of risks should be mitigated, transferred, retained, or avoided?
How should the use of financial instruments, such as derivatives or insurance,
be constrained and monitored to avoid creating new enterprise-level risks?

\paragraph{5. Integration into strategy and capital allocation}

Explain how ERM should be embedded into strategic planning, budgeting, and
capital allocation.
How would your design ensure that risk considerations influence major
investment decisions without becoming a purely formal requirement?

\paragraph{6. Reporting, dashboards, and escalation}

Propose principles for ERM reporting.
What information should be escalated, to whom, and under what conditions?
How would you design dashboards so that they support action rather than
reassurance?

\paragraph{7. Failure modes}

Identify two plausible scenarios in which your ERM design could still fail.
Explain which assumptions would break down and which parts of the system would
be most vulnerable. What early warning signals would you expect to see before
failure becomes unavoidable?

\subsection{Expected output}

Your group should produce a short written proposal (2--3 pages) summarizing
your ERM design and the reasoning behind it. There is no single correct answer;
the quality of your proposal will be assessed based on coherence, realism, and
ability to anticipate trade-offs and failure modes.



\end{document}
